Very simplified likelihood

### 2024-02-10 inst/very_simplified_likelihood.Rmd
consistentification

f <- function(a,r,n,log=FALSE){ # vectorized in 'a', treats 'r' as a vector of independent observations

  if(log){
    out <- 0
    for(rr in r){out <- out + f_vec_a(a,rr,n,log=TRUE)}
  } else {
    out <- 1
    for(rr in r){out <- out * f_vec_a(a,rr,n,log=FALSE)}
  }
  return(out)
}

MLE <- function(r,n){
    d <- 1e-6
   optimize(f,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
}
MLE(0,9)
MLE(3,9)
MLE(3,7)
```

@@

```{r}
a <- 0.23423434
b <- 1-a
B <- 1/b
n <- 9
r <- 3
c(
b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
(B-1)/prod(B+(5:8)),
f(a,r,n,log=FALSE))
```

@@

```{r,cache=TRUE}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 7
jj <- f(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n')
grid()
for(r in seq(from=0,to=n)){
  y <- f_vec_a(a,r,n,log=FALSE)
  if(r==0){y[length(y)] <- 1}  
  y <- y/max(y,na.rm=TRUE)
  if(r>0){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
  segments(x0=r/(n),y0=0.9,y1=1)
  text(x=r/(n),y=0.8,r)
}
}
plotter()
pdf(file="ninelikes.pdf")
plotter()
dev.off()
```

@@

```{r maxvscheco}
a <- read.table("formula1_2023.txt",header=TRUE)
a <- a[,seq_len(ncol(a)-1)]
d1 <- as.numeric(as.matrix(a)["Ocon",])
d2 <- as.numeric(as.matrix(a)["Gasly",])
d1[is.na(d1)] <- 22
d2[is.na(d2)] <- 22
d1
d2
MLE(d1,22)
MLE(d2,22)
```

@@

```{r 2samptest}
both <- c(d1,d2)
bothmax <- MLE(both,22)$maximum

d1max <- MLE(d1,22)$maximum
d2max <- MLE(d2,22)$maximum
f(bothmax,d1,22,log=TRUE)
f(bothmax,d2,22,log=TRUE)
f(d1max,d1,22,log=TRUE)
f(d2max,d2,22,log=TRUE)

Lambda <- (f(d2max,d1,22,log=TRUE)+f(d2max,d2,22,log=TRUE)) -(f(bothmax,d1,22,log=TRUE)+f(bothmax,d2,22,log=TRUE))
Lambda
pchisq(-2*Lambda,df=1,lower.tail=FALSE)
```

@@

  d1max <- MLE(d1,n)$maximum
  d2max <- MLE(d2,n)$maximum
  Lambda <- (f(d2max,d1,n,log=TRUE)+f(d2max,d2,n,log=TRUE)) -(f(bothmax,d1,n,log=TRUE)+f(bothmax,d2,n,log=TRUE))
  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
}

@@

## Masterchef

Look at the dotchart in the _MasterChef_ part of the `hyper2`
vignette.  We could use a Very Simplified Likelihood to estimate, say,
Laura's BT strength assuming that the other competitors are clones.

>>>>

f_vec_arn <- function(a,r,n,log=FALSE){ ## vectorized in 'a'; treats 'r' and 'n' as
                                ## vectors of independent observations

  M <- cbind(r,n)

  if(log){
    out <- 0
    for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,r=M[i,1],n=M[i,2],log=TRUE)}
  } else {
    out <- 1
    for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,r=M[i,1],n=M[i,2],log=FALSE)}
  }
  return(out)
}

MLE <- function(r,n){
    d <- 1e-6
    optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
}
MLE(0,9)
MLE(3,9)
MLE(3,7)
```

@@

```{r}
a <- 0.23423434
b <- 1-a
B <- 1/b
n <- 9
r <- 3
c(
b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
(B-1)/prod(B+(5:8)),
f_vec_arn(a,r,n,log=FALSE))
```

@@

```{r,label=likevis,cache=TRUE}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 7
jj <- f(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n')
grid()
for(r in seq(from=0,to=n)){
  y <- f_vec_a(a,r,n,log=FALSE)
  if(r==0){y[length(y)] <- 1}  
  y <- y/max(y,na.rm=TRUE)
  if(r>0){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
  segments(x0=r/(n),y0=0.9,y1=1)
  text(x=r/(n),y=0.8,r)
}
}
plotter()
pdf(file="ninelikes.pdf")
plotter()
dev.off()
```

@@

```{r maxvscheco}
a <- read.table("formula1_2023.txt",header=TRUE)
a <- a[,seq_len(ncol(a)-1)]
d_ocon  <- as.numeric(as.matrix(a)["Ocon",])
d_gasly <- as.numeric(as.matrix(a)["Gasly",])
d_ocon [is.na(d_ocon )] <- 22
d_gasly[is.na(d_gasly)] <- 22
d_ocon
d_gasly
MLE(d_ocon ,22)
MLE(d_gasly,22)
```

@@

```{r 2samptest}
both <- c(d_ocon,d_gasly)
bothmax <- MLE(both,22)$maximum

d_ocon_max  <- MLE(d_ocon ,22)$maximum
d_gasly_max <- MLE(d_gasly,22)$maximum
f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
f_vec_arn(bothmax,d_gasly,22,log=TRUE)
f_vec_arn(d_ocon_max ,d_ocon ,22,log=TRUE)
f_vec_arn(d_gasly_max,d_gasly,22,log=TRUE)

Lambda <- (
   (f_vec_arn(d_gasly_max  ,d_ocon,22,log=TRUE)+f_vec_arn(d_gasly_max  ,d_gasly,22,log=TRUE)) -
   (f_vec_arn(bothmax      ,d_ocon,22,log=TRUE)+f_vec_arn(bothmax      ,d_gasly,22,log=TRUE))
)
Lambda
pchisq(-2*Lambda,df=1,lower.tail=FALSE)
```

@@

  d1max <- MLE(d1,n)$maximum
  d2max <- MLE(d2,n)$maximum
  Lambda <- (
  (f_vec_arn(d2max  ,d1,n,log=TRUE)+f_vec_arn(d2max  ,d2,n,log=TRUE)) -
  (f_vec_arn(bothmax,d1,n,log=TRUE)+f_vec_arn(bothmax,d2,n,log=TRUE))
)
  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
}

@@

## Parkrun

```{r readresults}
results <- read.table("parkrun_results.txt",header=TRUE)
results
```

Above we see three parkrun results.  In parkrun number 148, I placed
229 out of a field of 318.

```{r definelikelihood}

L <- function(a,results){
   out <- a*0
   for(i in seq_len(nrow(results))){
   jj <- f_vec_a(a,r=results$place[i]-1,results$runners[i]-1,log=TRUE)
   out <- out + jj
   }
   out[is.infinite(out)] <- NA
   return(out)
}
```

```{r calculatelikes,cache=TRUE}
a <- seq(from=0.15,to=0.5,by=0.01)
La <- L(a,results)
La <- La-max(La,na.rm=TRUE)
La2 <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
La2 <- La2 - max(La2,na.rm=TRUE)
```

```{r plotlike}
plot(a,La)
points(a,La2)
abline(h=c(0,-2))
```

```{r optimsinglelike}
optimize(function(a){L(a,results)},c(0.3,0.4),maximum=TRUE)
```

```{r splitresults}
supp2 <- function(vec,place,runners){
  a_stirling <- vec[1]
  a_auckland <- vec[2]
  out <- 0
  for(i in 1:3){
    out <- out + f_vec_a(a_stirling,r=place[i]-1,runners[i]-1,log=TRUE)
  }
  for(i in 4:10){
    out <- out + f_vec_a(a_auckland,r=place[i]-1,runners[i]-1,log=TRUE)
  }
  return(out)
}
```

```{r,label=calcbothlikes,cache=FALSE}
x <- seq(from=0.1,to=0.8,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2,results$place,results$runners)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

```{r,label=plotbothlikes}
par(pty='s')
contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5))
abline(0,1)
```

```{r, label=onedimopt}
f <- function(a){supp2(c(a,a),results$place,results$runners)}
a <- seq(from=0.1,to=0.6,by=0.05)
alike <- sapply(a,f)
plot(a,alike,type='b')
optimize(f,c(0.1,0.6),maximum=TRUE)
```

```{r, label=twodimopt}
f2 <- function(v){supp2(v,results$place,results$runners)}
optim(par=c(0.6,0.3),f2,control=list(fnscale=-1))
```

```{r,label=showchisquareasymptotic}
pchisq(2*(7381.496-7380.314),df=1,lower.tail=FALSE)
```

## Education

The focal competitor

```{r}
rank <- c( 9, 7, 2, 3,2 ,8 ,5 ,4 ,9 )
class_size <- c(12, 17,23, 9,13,14,13,12,15)
course <- c("rings and modules","group theory","calculus","linear algebra",
"differential equations","topology","special relativity","fluid mechanics","Lie algebra")
category=c("pure","pure","applied","applied","applied","pure","applied","applied","pure")
data.frame(course,category,rank,class_size)
```

```{r}
a <- seq(from=0.1,by=0.01,to=0.7)
wp <- category=='pure'
wa <- category=='applied'
```

Pure first

```{r}
Lpure <- f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)
Lpure <- Lpure - max(Lpure,na.rm=TRUE)
plot(a,Lpure)
optimize(function(a){f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
optimize(function(a){f_vec_arn(a,rank[wa]-1,class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
```

```{r}
supp2_ed <- function(vec,place1,place2,runners1,runners2){
  out <- 0
  M1 <- cbind(place1,runners1)
  M2 <- cbind(place2,runners2)
  for(i in seq_len(nrow(M1))){
    out <- out + f_vec_arn(vec[1],r=M1[i,1]-1,n=M1[i,2]-1,log=TRUE)
  }
  for(i in seq_len(nrow(M2))){
    out <- out + f_vec_arn(vec[2],r=M2[i,1]-1,n=M2[i,2]-1,log=TRUE)
  }
  return(out)
}
```

```{r,cache=TRUE}
x <- seq(from=0.1,to=0.95,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,supp2_ed,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

```{r}
par(pty='s')
contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),
xlab='pure strength',ylab='applied strength')
abline(0,1)
```

```{r}
jj1 <- 
optim(
   c(.4,.8),
   function(vec){
      supp2_ed(vec,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
   control=list(fnscale = -1)
)
jj1
```

```{r}
jj2 <-
optimize(function(v){supp2_ed(c(v,v),rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
c(0.1,0.9),maximum=TRUE)
jj2
```

```{r}
pchisq(2*(jj1$value-jj2$objective),lower.tail=FALSE,df=1)
```
